Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions

### 2015-02-05 HGmodelMANUSCRIPT/HGmodel_draft3.tex
Added modified PN-vector-vs-hydration plot
updating information from blog to the manuscript

\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulation model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. In this work we compare the C--H bond vector order
parameters for glycerol backbone and choline headgroup between 14 different atomistic resolution
models and experiments in a fully hydrated lipid bilayer. The current models are not accurately enough to resolve the structure.
However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
would resolve the structure. Despite of the inaccuracy in the fully hydrated
structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been, and continues to be, progressed and discussed through the blog: \texttt{nmrlipids.blogspot.fi}. 
Everyone is invited to join the discussion and make contributions through the blog. 
The manuscript will be eventually submitted to an appropriate scientific journal. 
Everyone who has contributed to the work through the blog will be offered 
coauthorship. For more details see: \texttt{nmrlipids.blogspot.fi}.

@@

This work has been done as an open collaboration by using \texttt{nmrlipids.blogspot.fi} as a communication platform.
The approach is inspired by the Polymath project~\cite{gowers09}, however there are some essential differences. 
The manuscript pointing out problems in the glycerol backbone and
headgroup structure in the most used molecular dynamics simulation model for lipid bilayers (Berger based models) was used as a starting point~\cite{ollila13}.
After publishing the initial manuscript, the blog was opened for further contributions and discussion from anyone interested. 
All the contributions were done publicly through the blog and the blog contributors were offered coauthorship. The final author list
is based on the self-assesment of the authors' scientific contribution to the project.

Large portion of the used simulation files and scripts are shared through the GitHub organization \texttt{https://github.com/NMRlipids}
and the Zenodo community \texttt{https://zenodo.org/collection/user-nmrlipids}.

\subsection{Analysis}
\todo{ We should mention the sign issue and refer to the blog post:
http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html somewhere.
Maybe we should change the name of this section II.B. to: "Analysis and terminology of order parameters from experiments 
and simulations" or similar, and also move the "forking" explanation here.}
The order parameter of a hydrocarbon C--H vector is defined as $S_{{\rm CH}}=\frac{3}{2}\langle \cos^2 \theta-1 \rangle$, where
the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The order parameters from the measured dipolar splitting $d_\mathrm{CH}$ ($^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see \texttt{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.}.

@@

\sout {Berger-POPC-07~\cite{ollila07a}}\todoi{Not included in the Fig. 2, to be removed?}          &   POPC & 128 & 7290  & 323  & 56 & 56  & ? & SI \\

@@

CHARMM36\cite{klauda10}      & POPC   & 128 &  5120    & 303 & 150 & 100  & ? \todoi{Permanent link in progress by Hubert Santuz}   & SI \\

@@

CHARMM36\cite{klauda10}\todoi{Not included in the Fig. 2. Should we remove or include?}        & DOPC   & 128 &  5120    & 303 & 150 & 100  & ?\todoi{Permanent link in progress by Hubert Santuz}  & SI \\

@@

@@

Poger\cite{poger10}             & DPPC  & 128 & 5841 & 323 & 2$\times$100 & 2$\times$50 & \cite{pogerFILES}\todoi{Permanent link in progress by Patrick Fuchs} & SI \\

@@

@@

Rabinovich et al.\cite{rabinovich14}\todoi{Alexander Luybartsev, let us know if you share the files.}  & POPC   &  128 & 3840  & 303 & 100 & 80  & ? & \cite{rabinovich14}  \\
H\"ogberg et al.\cite{hogberg08}\todoi{Alexander Luybartsev, let us know if you share the files.}  & DMPC   &  98 & 3840  & 303 & 75 & 50 & ? & \cite{hogberg08} \\
Ulmschneider\cite{Ulmschneider09}    & POPC  & 128 & 3328 & 310 & 100 & 50 & \cite{ulmschneiderFILES}$^*$ & SI \\

@@

\begin{table*}[htb]
\centering
\caption{Simulated lipid bilayers containing cholesterol. The simulation file data sets marked with $^*$ include also part of the trajectory.
}\label{systemsCHOL}
\begin{tabular}{c c c c c c c c c c}
%\hline
Force field & lipid  & \footnote{The number of lipid molecules}N$_{\rm l}$  & \footnote{number of cholesterol molecules}N$_{\rm chol}$   &  \footnote{The number of water molecules}N$_{\rm w}$ & \footnote{Simulation temperature}T (K)  & \footnote{The total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time frames used in the analysis}t$_{{\rm anal}}$ (ns)\\
\hline
Berger-POPC-07~\cite{ollila07a}/H\"oltje-CHOL-13~\cite{holtje01,ferreira13}          &   POPC &128 & 0 & 7290  & 298  & 270 & 240 & \cite{bergerFILESpopc}$^*$ & \cite{ferreira15} \\
                               &   POPC &120 & 8 & 7290   & 298  & 100 & 80 & \cite{bergerFILESpopc7chol}$^*$ & \cite{ferreira13} \\
                               &   POPC &110 & 18 & 8481  & 298  & 100 & 80 & \cite{bergerFILESpopc15chol}$^*$ & \cite{ferreira13}  \\
                               &   POPC &84 & 44 & 6794   & 298  & 100 & 80 & \cite{bergerFILESpopc34chol}$^*$ & \cite{ferreira13} \\
                               &   POPC &64 & 64 & 10314  & 298  & 100 & 80 & \cite{bergerFILESpopc50chol}$^*$ & \cite{ferreira13} \\
                               &   POPC &50 & 78 & 5782   & 298  & 100 & 80 & \cite{bergerFILESpopc60chol}$^*$ & \cite{ferreira13} \\
CHARMM36\cite{klauda10,lim12}\todoi{Hubert Santuz, these files are already in Gitbub, however to get doi citation it should be put to Zenodo}    & POPC   & 128&  0 & 5120  & 303  & 150 & 100 & ? & SI  \\
 \todoi{Hubert Santuz, these files are already in Gitbub, however to get doi citation it should be put to Zenodo}                                & POPC   & 100  & 24  &  4960   & 303 & 200 & 100 & ? & SI \\
  \todoi{Hubert Santuz, these files are already in Gitbub, however to get doi citation it should be put to Zenodo}                               & POPC   & 80  & 80  &  4496    & 303 & 200 & 100 & ? & SI \\
MacRog\cite{maciejewski14}     & POPC   & 128 & 0  & 6400   & 310 & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI \\ 
                          & POPC   & 114  & 14 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
                          & POPC   & 72   & 56 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
                             & POPC   & 64  & 64 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
                             & POPC   & 56   & 72 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
\end{tabular}
\end{table*}

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of dipalmitoylphosphatidylcholine (DPPC) has been successful, 
allowing the order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC)~\cite{hong95b,gross97,dvinskikh05a}, 
1,2-dioleoyl-sn-glycero-3-phosphocholine (DOPC)~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments.
The experimental absolute values of glycerol backbone and choline order parameters from various publications are shown in Fig.~\ref{HGorderparameters}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonSorted.pdf}
%\todo{DONE}\\
%The data in this figure should be updated to be consistent with the systems listed in Table~\ref{systems}. At least, Tjornhammar et al and CHARMM-UA values are missing. Some results in the figure are outdated due to longer simulations reported etc.} \\
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}} \\
\todo{Markus has now made a new figure with signs included. I think that this is very good. I also think that we should try to get the interactive figure behind permalink (i.e. Zenodo or figshare) and cite that in the paper.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

To get a general idea of the quality of the glycerol backbone and choline headgroup structures in different models, we calculated the absolute
values of the order parameters for these parts from fourteen different lipid models (Table~\ref{systems}) and 
plotted the results together with experimental values in Fig.~\ref{HGorderparameters}.
Two criteria were used to judge the quality of the model: there must not be significant  {\bf forking} in the $\alpha$ and $\beta$ carbons,
there must be only moderate forking in the g$_3$ carbon and there must be significant forking in the g$_1$ carbon, {\bf absolute values}
should be preferably inside to the subjective sweet spots determined from experiments (blue shaded regions in Fig.~\ref{HGorderparameters}).
None of the studied force fields fulfils these criteria. However, three force fields perform better than others, namely CHARMM36~\cite{klauda10}, MacRog~\cite{maciejewski14} and GAFFlipid~\cite{dickson12}.
The results for each force field in respect to the above criteria are summarized in Figure~\ref{FullHydrationComparisonTable}.
\todo{This kind of table should be done or not? DONE, see Fig.~\ref{FullHydrationComparisonTable}}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonTable.pdf}
  \caption{\label{FullHydrationComparisonTable}
  Ranking of force fields based on data of Fig.~\ref{HGorderparameters}.
  } 
\end{figure}

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured (see
\texttt{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
for more discussion.
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.}
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
showing increase for both segments with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone g$_3$- and g$_2$- order parameters were observed with dehydration, 
while g$_1$ remained practically unchanged~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/dehydration.pdf}
\todo{Several people think that this plot should be clarified. ------> DONE.}
  \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
  }
\end{figure}

@@

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration in Fig.~\ref{PNangle}.
However, the qualitative agreement in the lipid response to dehydration does not guarantee the correct 
free energy landscape if the simulation model has an incorrect structure. The influence of this issue to 
dehydration energetics studied with simulations~\cite{eun09,schneck12} is left for future studies.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{PNangles.eps}

>>>>

\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulation model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. In this work we compare the C--H bond vector order
parameters for glycerol backbone and choline headgroup between 14 different atomistic resolution
models and experiments in a fully hydrated lipid bilayer. The current models are not accurately enough to resolve the structure.
However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
would resolve the structure. Despite of the inaccuracy in the fully hydrated
structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been, and continues to be, progressed and discussed through the blog: \url{nmrlipids.blogspot.fi}. 
Everyone is invited to join the discussion and make contributions through the blog. 
The manuscript will be eventually submitted to an appropriate scientific journal. 
Everyone who has contributed to the work through the blog will be offered 
coauthorship. For more details see: \url{nmrlipids.blogspot.fi}.

@@

This work has been done as an open collaboration by using \url{nmrlipids.blogspot.fi} as a communication platform.
The approach is inspired by the Polymath project~\cite{gowers09}, however there are some essential differences. 
The manuscript pointing out problems in the glycerol backbone and
headgroup structure in the most used molecular dynamics simulation model for lipid bilayers (Berger based models) was used as a starting point~\cite{ollila13}.
After publishing the initial manuscript, the blog was opened for further contributions and discussion from anyone interested. 
All the contributions were done publicly through the blog and the blog contributors were offered coauthorship. The final author list
is based on the self-assesment of the authors' scientific contribution to the project.

Large portion of the used simulation files and scripts are shared through the GitHub organization \url{https://github.com/NMRlipids}
and the Zenodo community \url{https://zenodo.org/collection/user-nmrlipids}.

\subsection{Analysis}
\todo{ We should mention the sign issue and refer to the blog post:
http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html somewhere.
Maybe we should change the name of this section II.B. to: "Analysis and terminology of order parameters from experiments 
and simulations" or similar, and also move the "forking" explanation here.}
The order parameter of a hydrocarbon C--H vector is defined as $S_{{\rm CH}}=\frac{3}{2}\langle \cos^2 \theta-1 \rangle$, where
the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The order parameters from the measured dipolar splitting $d_\mathrm{CH}$ ($^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see \url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.}.

@@

%\sout {Berger-POPC-07~\cite{ollila07a}}\todoi{Not included in the Fig. 2, to be removed?}          &   POPC & 128 & 7290  & 323  & 56 & 56  & ? & SI \\
\todoi{DONE} 

@@

CHARMM36\cite{klauda10}      & POPC   & 128 &  5120    & 303 & 150 & 100  & \cite{charmm36files}$^*$ \todoi{DONE}   & SI \\

@@

%CHARMM36\cite{klauda10}\todoi{Not included in the Fig. 2. Should we remove or include?}        & DOPC   & 128 &  5120    & 303 & 150 & 100  & ?\todoi{Permanent link in progress by Hubert Santuz}  & SI \\
\todoi{DONE} \todoi{Permanent link for DOPC CHARMM36 simulation by Hubert Santuz is not required now. However, it would be probably useful in the future.}

@@

GAFFlipid\cite{dickson12}       & DPPC & 72  & 2197  & 323 & 90 & 50  & \cite{GAFFlipidFILESdppc}$^*$ & SI \\

@@

Poger\cite{poger10}             & DPPC  & 128 & 5841 & 323 & 2$\times$100 & 2$\times$50 & \cite{pogerFILESpme1,pogerFILESpme2}$^*$\todoi{DONE} & SI \\

@@

Slipid\cite{jambeck12}          & POPC & 128 & 5120 & 303 & 200 & 150 & \cite{slipidsFILESpopc}$^*$ & SI \\

@@

H\"ogberg08\cite{rabinovich14}\todoi{Alexander Luybartsev, let us know if you share the files.}  & POPC   &  128 & 3840  & 303 & 100 & 80  & ? & \cite{rabinovich14}  \\
H\"ogberg08\cite{hogberg08}\todoi{Alexander Luybartsev, let us know if you share the files.}  & DMPC   &  98 & 3840  & 303 & 75 & 50 & ? & \cite{hogberg08} \\
Ulmschneiders\cite{Ulmschneider09}    & POPC  & 128 & 3328 & 310 & 100 & 50 & \cite{ulmschneiderFILES}$^*$ & SI \\

@@

\begin{table*}[htb]
\centering
\caption{Simulated lipid bilayers containing cholesterol. The simulation file data sets marked with $^*$ include also part of the trajectory.
}\label{systemsCHOL}
\begin{tabular}{c c c c c c c c c c}
%\hline
Force field & lipid  & \footnote{The number of lipid molecules}N$_{\rm l}$  & \footnote{number of cholesterol molecules}N$_{\rm chol}$   &  \footnote{The number of water molecules}N$_{\rm w}$ & \footnote{Simulation temperature}T (K)  & \footnote{The total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time frames used in the analysis}t$_{{\rm anal}}$ (ns)\\
\hline
Berger-POPC-07~\cite{ollila07a}/H\"oltje-CHOL-13~\cite{holtje01,ferreira13}          &   POPC &128 & 0 & 7290  & 298  & 270 & 240 & \cite{bergerFILESpopc}$^*$ & \cite{ferreira15} \\
                               &   POPC &120 & 8 & 7290   & 298  & 100 & 80 & \cite{bergerFILESpopc7chol}$^*$ & \cite{ferreira13} \\
                               &   POPC &110 & 18 & 8481  & 298  & 100 & 80 & \cite{bergerFILESpopc15chol}$^*$ & \cite{ferreira13}  \\
                               &   POPC &84 & 44 & 6794   & 298  & 100 & 80 & \cite{bergerFILESpopc34chol}$^*$ & \cite{ferreira13} \\
                               &   POPC &64 & 64 & 10314  & 298  & 100 & 80 & \cite{bergerFILESpopc50chol}$^*$ & \cite{ferreira13} \\
                               &   POPC &50 & 78 & 5782   & 298  & 100 & 80 & \cite{bergerFILESpopc60chol}$^*$ & \cite{ferreira13} \\
CHARMM36\cite{klauda10,lim12}\todoi{DONE}    & POPC   & 128&  0 & 5120  & 303  & 150 & 100 & \cite{charmm36files}$^*$  & SI  \\
 \todoi{DONE}                                & POPC   & 100  & 24  &  4960   & 303 & 200 & 100 & \cite{charmm36files20perCHOL}$^*$ & SI \\
  \todoi{DONE}                               & POPC   & 80  & 80  &  4496    & 303 & 200 & 100 & \cite{charmm36files50perCHOL}$^*$ & SI \\
MacRog\cite{maciejewski14}     & POPC   & 128 & 0  & 6400   & 310 & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI \\ 
                          & POPC   & 114  & 14 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
                          & POPC   & 72   & 56 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
                             & POPC   & 64  & 64 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
                             & POPC   & 56   & 72 & 6400  & 310  & 400 & 200 & \cite{macrogCHOLfiles}$^*$ & SI    \\
\end{tabular}
\end{table*}

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of dipalmitoylphosphatidylcholine (DPPC) has been successful, 
allowing the order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC)~\cite{hong95b,gross97,dvinskikh05a}, 
1,2-dioleoyl-sn-glycero-3-phosphocholine (DOPC)~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments.
The experimental absolute values of glycerol backbone and choline order parameters from various publications are shown in Fig.~\ref{HGorderparameters}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonSorted.pdf}
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}} \\
\todo{I think that this figure is now very good. There is also the interactive version by Hubert Santuz now in https://plot.ly/~HubertSantuz/72/lipid-force-field-comparison/
we should figure out which is the most practical way to put that behind permalink once it finalized (Zenodo, figshare or something else?) and then put a citation in the paper.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

To get a general idea of the quality of the glycerol backbone and choline headgroup structures in different models, we calculated the absolute
values of the order parameters for these parts from fourteen different lipid models (Table~\ref{systems}) and 
plotted the results together with experimental values in Fig.~\ref{HGorderparameters}.
Two criteria were used to judge the quality of the model: there must not be significant  {\bf forking} in the $\alpha$ and $\beta$ carbons,
there must be only moderate forking in the g$_3$ carbon and there must be significant forking in the g$_1$ carbon, {\bf absolute values}
should be preferably inside to the subjective sweet spots determined from experiments (blue shaded regions in Fig.~\ref{HGorderparameters}).
None of the studied force fields fulfils these criteria. However, three force fields perform better than others, namely CHARMM36~\cite{klauda10}, MacRog~\cite{maciejewski14} and GAFFlipid~\cite{dickson12}.
The results for each force field in respect to the above criteria are summarized in Figure~\ref{FullHydrationComparisonTable}.
\todo{The table is now done, see Fig.~\ref{FullHydrationComparisonTable}. The text should be updated accordingly.}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonTable.pdf}
  \caption{\label{FullHydrationComparisonTable}
  Ranking of force fields based on data of Fig.~\ref{HGorderparameters}.
  } 
\end{figure}

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured (see
\url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
for more discussion.
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.}
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
showing increase for both segments with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone g$_3$- and g$_2$- order parameters were observed with dehydration, 
while g$_1$ remained practically unchanged~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/dehydration.pdf}
\todo{Figure has been modified to elucidate changes. The text should be updated accordingly. The same applies to the cholesterol figure.}
  \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
  }
\end{figure}

@@

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration in Fig.~\ref{PNangle}.
However, the qualitative agreement in the lipid response to dehydration does not guarantee the correct 
free energy landscape if the simulation model has an incorrect structure. The influence of this issue to 
dehydration energetics studied with simulations~\cite{eun09,schneck12} is left for future studies.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/dehydrationPN.pdf}
